Developing the algebraic hierarchy with type classes in Coq

### 2010-03-22 roughdiamond.tex
Rewrite first two pages in Eelis-style.

\documentclass{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage[utf8x]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs %i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{Developing the algebraic hierarchy with type classes in Coq}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Finally, we provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions.
\end{abstract}

\section{Introduction}
Developing formalized libraries for mathematics requires a detailed analysis of the architecture~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large software project. 
The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
inheritance: A ring is a type with two operations. How do we inherit from the two theories of
monoids? 2. Sharing of notations.

Several solutions have been proposed for the Coq theorem prover%~\cite{Coq,BC04}
: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded
\footnote{Modules are not first class, but currently they seem to be superior for extraction
purposes%~\cite{letouzey02}%
.}, considering
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

In this fashion we implement\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}}
 algebraic theories as type classes, not instances, thus addressing the
challenges above; see Section~\ref{classes}. Type classes have recently entered Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq
type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.

@@

We abstract from concrete algebra and formalize many-sorted universal algebra, including a construction of
free $\Sigma$-algebras; Section~\ref{univ}. We intend to use the current work as a basis for
a library for computational analysis.  Thus we want to be able to replace data structures
with more efficient ones. For instance, Coq provides unary, binary and machine numbers. We
use ad-hoc polymorphism with type classes to develop a unified library for them. The interfaces for
$\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see
Section~\ref{interfaces}.

@@

\section{Algebraic hierarchy using type classes}\label{classes}
\vspace*{-2ex}
\paragraph{Canonical names.}\label{names}
Sharing of notations is an important aspect of the design of a mathematical library. Haskell uses
type classes to overload notations. In Coq we assign a class to every operation, thus avoiding Coq's notation |scope| mechanism.
 Each such operation class has only \emph{one} constructor, hence no record is constructed.
\begin{code}
 Class Equiv A := equiv: relation A.
 Infix "==" := equiv (at level 70, no associativity).
\end{code}
% Instance: Params (@equiv) 2.
\vspace*{-5ex}
\paragraph{Hierarchy.}\label{hierarchy}
In our formalization of algebraic structures we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that |op| does not depend on the proof that |e| is an equivalence. We use Coq's implicit
quantification (|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem
and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
theorem. Both features are new in Coq and stem from the type class implementation.

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively harmless: most
terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
as in the definition of a semiring: a type with two monoid structures on it. We avoid some
of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
properties, although the properties of a semiring are not structurally included in the ring
properties.

Manifest fields in records allow us to fix a field of a dependent record.
Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes would
be awkward due to problems with equality. We avoided these problems by shifting the fields
from the dependent record to its argument, thus mimicking manifest fields by using type classes.
Since |op| is \emph{instantiated} with |mult|, we need not worry about the
equality |op=mult|.

>>>>

\documentclass{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage[utf8x]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs %i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{Developing the algebraic hierarchy with type classes in Coq}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its newly obtained type class implementation to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties affecting efficient computation are reduced to a bare minimum.
\end{abstract}

\section{Introduction}
The development of libraries for formalized mathematics presents many software engineering challenges ~\cite{C-corn,DBLP:conf/types/HaftmannW08} arising from the fact that it is far from obvious how the clean, idealized concepts from everyday mathematics should be represented using the facilitities provided by concrete theorem provers and their formalisms, in a way that is both mathematically faithful and convenient to work with.

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation.

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's recently obtained type classes, and show how their features together with a key design pattern let us effectively address the challenges mentioned above.

Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation. In particular, we want to be able to effortlessly swap implementations of number representations. Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.

To express these elegantly and without duplication, our development\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of the basics of category theory and multi-sorted universal algebra, all expressed using type classes where appropriate.

@@

@@

\section{Algebraic hierarchy using type classes}\label{classes}
\vspace*{-2ex}
We represent each structure in the algebraic hierarchy as a type class. In Coq, type classes and their instances are realized as record types (``dictionaries'') and registered constants of these types. This immediately leads to the familiar question of which components of the structure are to become parameters, and which are to become record(/class) fields. By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e. carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields. Type classes defined this way are essentially predicates with automatically resolved proofs.

Conventional wisdom warns that while this approach is theoretically very flexible (as we will see momentarily), one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).

These are legitimate concerns that we avoid by exploiting the way Coq type classes and their support infrastructure work, using \emph{operational type classes}: classes with a single field representing a single relation or operation in isolation. Such classes are treated specially by Coq in being translated to mere definitions rather than records, with the field projection becoming the identity function.
\begin{code}
 Class Equiv A := equiv: relation A.
 Infix "=" := equiv (at level 70, no associativity).
\end{code}
These operational type classes serve to establish \emph{canonical names}, which not only let us bind notations, but also allow us to make their declaration and passing implicit in most contexts. For instance, using the following definition of semirings, all structural parameters (represented by operational classes declared with curly brackets) will be implicitly resolved by the type class mechanism rather than listed explicitly whenever we talk about semirings.
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
The concrete Coq infrastructure (added specifically to support type classes) that makes this work consists of implicit quantification (when declaring a semiring), and maximally inserted implicit arguments (when stating that something is a semiring, and when referring to operations and relations).

Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.

First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions \cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.

Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths (obviating the need for things like the proposed coercion pullbacks~\cite{Hints}).

Third, since our structural type classes are mere (proof-irrelevant) predicates, overlap between their instances is a non-issue. Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses.

% Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
% as in the definition of a semiring: a type with two monoid structures on it. We avoid some
% of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
% properties, although the properties of a semiring are not structurally included in the ring
% properties.
